Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-29 text/ms.tex
Edit edit edit

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology was the asteroid marking the mass extinction of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value (GEV) distribution has
been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
the distribution of the most extreme event per time interval (e.g.~heaviest
rainfall per year), but this approach requires long time series that can be
condensed into time intervals. (2) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis did not quantify the probability of black
swans on a continuous scale or allow for population dynamics. (3) Fitting
a catastrophic mixture distribution state space model allows one to quantify
the probability that population events are extreme on a continuous scale but is
data intensive \citep{ward2007}.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the Gompertz
model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

We extended the standard Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

@@

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimates of $\nu$. We formally investigated
these relationships by fitting beta regression multilevel models to the
probability that $\nu$ was less than $10$ (i.e.~approximately the probability
of heavy tails). For covariates that were derived from Gompertz model parameter
estimates, we incorporated standard deviations around the means. To account for
broad patterns of phylogenetic relatedness, we allowed for varying intercepts
at the taxonomic class, order, and species level (Supporting Material).

@@

We found strong but rare evidence for black-swan dynamics. Defining black-swan
dynamics as a greater than $0.5$ probability that $\nu < 10$, we observed black
swans most frequently for birds (\birdPH\%) followed by mammals (\mammalsPH\%),
insects (\insectsPH\%), and fishes (\fishPH\%) (Fig.~\ref{fig:nu-coefs},
\ref{fig:heavy-ts}). Black swans were taxonomically widespread, occurring for
at least one population in \POrdersHeavy\% of the taxonomic orders recorded.
Accounting for time series length and partially pooling inference across
taxonomic class and order with a multilevel model, there was stronger evidence
for black swans in insect populations than is apparent in
Fig.~\ref{fig:nu-coefs} --- four of 10 orders with the highest median
probability of heavy tails were insect orders --- however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $>10$ in
$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations although the majority of
$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
similarly observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
in a population with 60 time steps than in one with 30 time steps
(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
increased time series length was small ($0.\pHeavyNSixty$ vs.\
$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the UK was thought to reduce population productivity and
make the population vulnerable when a red tide hit in 1968 \citep{potts1980}.
The population experienced a black swan downswing, but the new availability of
nest sites resulted in a rapid population upswing \citep{potts1980}. As
another example, an interaction between environmentally and predation-mediated
population cycles are thought to have caused a downward black swan for a water
vole population \citep{saucy1994}. Other black swans were the result of
a sequence of extreme climate events on their own. For example, severe winters
in 1929, 1940--1942, and 1962--1963 were associated with black swan downswings
in grey heron abundance in the UK \citep{stafford1971}. Our analysis finds
that the last event was a combination of two black swan events in a row and it
took the population three times longer to recover than predicted
\citep{stafford1971}.

@@

We found strong evidence that the process noise in ecological time series was
heavy-tailed under the Gompertz or Ricker-logistic models in
\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
swans more frequently for longer time series and for populations with a smaller
magnitude of process error, as we might expect. However, we failed to find
strong evidence that black swans were associated with density dependence,
population growth rate, or lifespan. In cases that we verified, black swans
were often a result of the interaction between elements of extreme climate,
predators and parasite cycles, and strong changes in human pressures. Our
empirical results, sensitivity analyses, and simulation tests suggest that
estimating the tail shape of process noise is a viable method of detecting
black swan population dynamics and if anything will underestimate the
probability of black swans. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
\citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the lognormal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black swans
in a large proportion of populations. Furthermore, by probabilistically
quantifying process noise tails on a continuous spectrum, and allowing for
non-stationary time series and density dependence, our analysis allows for
a more nuanced description of the evidence of ecological black swans. In an
earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
bird population abundance. However, this finding was challenged by
\citet{allen2001}, who showed that mixing data across species could create
falsely generate heavy tails. This rebuttal supports our use of the
t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with
inverse-gamma-distributed variances,][]{gelman2014}.

@@

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point; it is likely
that some of the black swans in our dataset are recording mistakes mistakes.
Conversely, recording mistakes could mask heavy tails if initial observations
are discarded or altered as assumed errors. Third, the temporal scales of
observation and population dynamics vary considerably across populations in the
GPDD and this likely influences the detection of heavy tails. For example, if
we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss black
swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Can we isolate the components of
ecological dynamics that experience black swans events by moving from
phenomenological models such as the Gompertz to mechanistic models, that take
into account, for example, recruitment? We expect that greater insight into the
mechanisms and covariates of ecological black swans may be best obtained
through specific geographic and taxonomic subsets of data where longer time
series are available \citep[e.g.][]{segura2013}. Finally, what is the impact of
allowing for black swans in forecasts of ecological risk? Financial analysts
are concerned with the shape of financial-return-distribution tails because
these impact estimates of risk --- the probability of a specific magnitude of
undesired event occurring \citep{rachev2008}. It is possible that a comparable
focus in ecology would alter our perception of ecological forecasts such as
extinction risk.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF). A similar focus could benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and tune future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value less than -5 is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero ($2.9\cdot10^{-5}$\%). (c--e) Simulated population
dynamics from a Gompertz model with process noise drawn from t distributions
with different values of $\nu$. Coloured dots in panels c and d represent jumps
with less than a 1 in 1000 chance of occurring in a normal distribution. (f--h)
Estimates of $\nu$ from models fit to the times series in panels c--e. Shown
are the posterior samples (histograms), median and interquartile range of the
posterior (IQR) (dots and line segments), and the exponential prior on $\nu$
(dashed lines). Colour shading behind panels f--h illustrates the region of
heavy tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of
heavy tails) for populations arranged by taxonomic order and sorted by
decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red to yellow points highlight
populations with a high probability of heavy-tailed dynamics.
} \label{fig:nu-coefs} \end{center} \end{figure}

>>>>

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology is the asteroid marking the mass extinctions of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, storm severity, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value distribution has been
applied to environmental data \citep[e.g.][]{katz2005}. This distribution
describes the most extreme event per time interval (e.g.~heaviest rainfall per
year), but requires long time series that can be condensed into time intervals.
(2) Fitting a catastrophic mixture distribution in a state space model allows
one to quantify the probability that population events are extreme but is also
data intensive \citep{ward2007} (3) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis did not quantify the probability of black
swans on a continuous scale or allow for standard population dynamics.

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:

\vspace{-6mm}
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

We extended the Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:

\vspace{-6mm}
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

@@

We investigated possible covariates of heavy-tailed population dynamics
visually and through multilevel modelling. We plotted characteristics of the
time series ($\sigma$, $\lambda$, $b$, and time-series length) along with two
life-history characteristics (body length and maximum lifespan obtained from
\citet{brook2006a}) against our estimated probability of heavy tails, Pr$(\nu
< 10) > 0.5$. We formally investigated these relationships by fitting beta
regression multilevel models. For covariates that were derived from Gompertz
model parameter estimates, we incorporated standard deviations around the
means. To account for broad patterns of phylogenetic relatedness, we allowed
for varying intercepts at the taxonomic class, order, and species level
(Supporting Material).

@@

We found strong but rare evidence for black-swan population dynamics. Defining
black-swan dynamics as a greater than $0.5$ probability that $\nu < 10$, our
main Gompertz model found evidence for heavy tails most frequently for birds
(\birdPH\%) followed by mammals (\mammalsPH\%), insects (\insectsPH\%), and
fishes (\fishPH\%) (Figs~\ref{fig:nu-coefs}, \ref{fig:heavy-ts}). Black swans
were taxonomically widespread, occurring for at least one population in
\POrdersHeavy\% of the taxonomic orders recorded. Accounting for time series
length and partially pooling inference across taxonomic class and order with
a multilevel model, there was stronger evidence for black swans in insect
populations than is apparent in Fig.~\ref{fig:nu-coefs} --- four of 10 orders
with the highest median probability of heavy tails were insect orders ---
however, there was considerable uncertainty in these estimates
(Fig.~\ref{fig:order-estimates}).

The majority of our heavy-tailed estimates were robust to alternative
population models and observation error (Fig.~\ref{fig:alt}). Including an
autocorrelation structure in the residuals, modelling population growth rates
without density dependence, or modelling the population dynamics as
Ricker-logistic did not systematically alter our conclusions. Allowing for
a fixed quantity of observation error decreased our estimates of black-swan
dynamics, increasing the median estimate of $\nu$ from $<10$ to $\ge 10$ in
$\baseNuTenObsTenSwitch$ of $\baseNuTen$ populations, although the majority of
$\nu$ estimates remained qualitatively similar (Fig.~\ref{fig:alt})

Across populations, the probability of observing black swan dynamics was
positively related to time-series length and negatively related to magnitude of
process noise ($\sigma$) but not clearly related to population growth rate
($\lambda$), density dependence ($b$), or maximum lifespan
(Fig.~\ref{fig:correlates}, \ref{fig:correlate-coefs}). Longer time-series
length was the strongest covariate of observing black swan dynamics (as was
also observed in simulation testing, Fig.~\ref{fig:sim-nu}). For example,
we were about \pIncHeavyNThirtyNSixty~times more likely to observe heavy tails
in a population with 60 time steps than in one with 30 time steps
(Fig.~\ref{fig:perc-inc-p}). However, the absolute change in probability with
increased time series length was small ($0.\pHeavyNSixty$ vs.\
$0.\pHeavyNThirty$ in the previous example, Fig.~\ref{fig:correlates}).

We assembled the documented causes of black swan population dynamics in seven
populations with high probabilities of heavy tails (Table 1). The majority of
documented events were downward black swans and the majority involved
a confluence of events. For example, a shortage of nest sites for the shag
long-necked bird in the United Kingdom was thought to reduce population
productivity and make the population vulnerable when a red tide hit in 1968
\citep{potts1980}. The population experienced a black swan downswing, but the
new availability of nest sites resulted in a rapid population upswing
\citep{potts1980}. As another example, an interaction between environmental-
and predation-mediated population cycles are thought to have caused a downward
black swan for a water vole population \citep{saucy1994}. Other black swans
were the result of a sequence of extreme climate events on their own. For
example, severe winters in 1929, 1940--1942, and 1962--1963 were associated
with black swan downswings in grey heron abundance in the United Kingdom
\citep{stafford1971}. Our analysis finds that the last event was a combination
of two black swan events in a row and it took the population three times longer
to recover than predicted \citep{stafford1971}.

@@

We found strong evidence that the process noise in ecological time series was
heavy-tailed under  Gompertz or Ricker-logistic models in
\overallMinPerc--\overallMaxPerc\% of populations assessed. We detected black
swans more frequently for longer time series and for populations with a smaller
magnitude of process error, as we might expect. However, we failed to find
strong evidence that black swans were associated with density dependence,
population growth rate, or lifespan. In cases that we verified, black swans
were often a result of the interaction between elements of extreme climate,
predators and parasite cycles, and strong changes in human pressures. Our
empirical results, sensitivity analyses, and simulation tests suggest that
estimating the tail shape of process noise is a viable method of detecting
black swan population dynamics and if anything will underestimate the
probability of black swans. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics --- particularly as the
frequency and magnitude of climatic extremes increase over the next century
\citep{easterling2000,ipcc2012}.

Our results clarify previous related analyses. Previous work with an older
version of the GPDD assessed the distribution of abundance time series but
focused on identifying if the log-normal distribution was the most frequently
parsimonious model \citep{halley2002}. For heavy-tailed distributions,
\citet{halley2002} fit the extremely heavy-tailed Cauchy distribution and the
four-parameter Levy stable distribution and found little information criteria
support for these distributions in longer time series. However, black swan
events are by definition rare, and the majority of time series in the GPDD are
short for these purposes. Therefore, we would not expect to observe black swans
in a large proportion of populations. Furthermore, by probabilistically
quantifying process noise tails on a continuous spectrum, and allowing for
non-stationary time series and density dependence, our analysis allows for
a more nuanced description of the evidence of ecological black swans. In an
earlier study, \citet{keitt1998} described heavy (power law) tails in breeding
bird population abundance. However, this finding was challenged by
\citet{allen2001}, who showed that mixing data across species could
falsely generate heavy tails. This rebuttal supports our use of the
t distribution to represent heavy tails, since a t distribution can be
represented by a mixture of normal distributions \citep[with the same mean and
inverse-gamma-distributed variances,][]{gelman2014}.

@@

There are a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations --- the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find qualitatively similar evidence for black swans in many other large
taxonomic or
geographic samples of populations, the common forces driving
those black swans likely differ (e.g.~severe
winters in Table~1). Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point; it is likely
that some of the black swans in our dataset are recording mistakes.
Conversely, recording mistakes could mask heavy tails if extreme observations
are discarded or altered as assumed errors. Third, the temporal scales of
observation and population dynamics vary considerably across populations in the
GPDD and this likely influences the detection of heavy tails. For example, if
we make frequent observations relative to generation time (e.g.~for many
large-bodied mammals) we will average across generations and perhaps miss black
swans. Furthermore, if we census populations infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen black swan dynamics.

Our results suggest important future issues to address about ecological black
swans. Given that black swans do occur, can we forecast the probability of
black swans in space and time? Furthermore, what management policies allow us
to detect them quickly after they happen? Can we isolate the components of
ecological dynamics that experience black swans by moving from
phenomenological models such as the Gompertz to mechanistic models, that take
into account, for example, recruitment? We expect that greater insight into the
mechanisms and covariates of ecological black swans may be best obtained
through specific geographic and taxonomic subsets of data where longer time
series with low levels of observation error are available
\citep[e.g.][]{segura2013}. Finally, what is the impact of
allowing for black swans in forecasts of ecological risk? Financial analysts
are concerned with the shape of financial-return-distribution tails because
these impact estimates of risk --- the probability of a specific magnitude of
undesired event occurring \citep{rachev2008}. It is possible that a comparable
focus in ecology would alter our perception of ecological forecasts such as
extinction risk.

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/t-nu-eg.pdf}
\caption{
An illustration of fitting population dynamic models that allow for heavy
tails, represented by the Student-t degrees of freedom parameter $\nu$. (a, b)
The probability density for t distributions with a scale parameter of 1 and
different values of $\nu$. Small values of $\nu$ create heavy tails. As $\nu$
approaches infinity the distribution approaches the normal distribution. For
example, at $\nu = 2$, the probability of drawing a value more than five
standard deviations below the mean is
1.8\%, whereas the probability of drawing such a value from a normal
distribution is nearly zero. (c--e) Simulated population dynamics from
a Gompertz model with process noise drawn from t distributions with different
values of $\nu$. Coloured dots in panels c and d represent jumps with less than
a 1 in 1000 chance of occurring in a normal distribution. (f--h) Estimates of
$\nu$ from models fit to the times series in panels c--e. Shown are the
posterior samples (histograms), median and interquartile range of the posterior
(IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed
lines). Colour shading behind panels f--h illustrates the region of heavy
tails.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.44\textwidth]{../analysis/nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\
populations of birds, mammals, insects, and fishes. Small values of $\nu$
($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest
approximately normal-tailed dynamics. Vertical points and line segments
represent posterior medians and 50\% / 90\% credible intervals for individual
populations. Inset plots show probability that $\nu < 10$ (probability of
heavy tails) for populations arranged by taxonomic order and sorted by
decreasing mean Pr($\nu < 10$). Taxonomic orders with three or fewer
populations in panel a are omitted for space. Red to yellow points highlight
populations with a high to moderately high probability of heavy-tailed dynamics.
} \label{fig:nu-coefs} \end{center} \end{figure}
